Gradual Soundness: Lessons from Static Python

### 2022-04-23 pj-7.1/paper.tex
p edits keep coming

%% TODO after submission, before camera-ready
%% - reword 'fields' to attribute/descriptor/property?

@@

Generic types are more difficult to enforce because a tag by itself does not
describe their contents.
For example, suppose that a function expects inputs that match
the type \code{Dict[int,int]}; that is, dictionaries with integer keys and
integer values.
A constant-time check for Python dictionaries is not enough to protect the
function against dicts with ill-typed elements.
The \SP{} solution is to provide a second kind of dictionary that ensures
the validity of its elements.
These different values have different types:
\begin{enumerate}
  \item
    The \emph{shallow} type $\sptrawpydict$ gets enforced with a constant-time check
    for dict values that ignores elements.
    Consequently, a parameterized surface type such as $\sptpydict{K}{V}$
    does not promise anything about keys and values.
  \item
    The \emph{concrete} type $\sptchkdict{K}{V}$ gets enforced with a constant-time,
    compound tag check for concrete dict values that were initialized with
    exactly the same key and value types.
    %% Programmers must use a type-annotated contructor to make a concrete-typed checked dict.
\end{enumerate}
%
These options let programmers decide whether to edit their code to support concrete types
or leave it as-is with shallow types.

Often, the shallow types are more attractive because concrete types impose
nonlocal changes.
If one type changes to concrete, then several changes may need to happen:
all values that reach this type must be initialized as concrete,
all typed clients of these values must expect concrete values,
all values that reach those clients must be concrete,
and so on and so forth.

@@

\begin{table}[t]
  \caption{Worst-case costs for shallow and concrete types (T=typed, U=untyped)}
  \label{t:shallow-vs-concrete}

  \centering
  \(\begin{array}{l@{}rr@{\qquad}rr@{~~}@{\qquad}r@{~~}r}
                   & \mbox{Constructor} & \mbox{Cast} & \mbox{T-Read} & \mbox{T-Write} & \mbox{U-Read} & \mbox{U-Write} \\\hline
    \sptrawpydict         &           0 &        O(1) &          O(1) &              0 &             0 &              0 \\
    \sptchkdict{K}{V}     &        O(N) &        O(1) &             0 &              0 &             0 &           O(1)
  \end{array}\)
\end{table}

@@

Gradual class hierarchies are rarely studied in the literature, especially
for a language where truly untyped classes can enter the mix.
Thorn~\cite{wzlov-popl-2010} and SafeTS~\cite{rsfbv-popl-2015}, for example,
provide separate hierarchies for untyped and (gradually) typed classes.
The closest related work, for Nom~\cite{mt-oopsla-2017}, implements
a rather flexible point in the design space.
\SP{} implements a simpler design that restricts types and
inheritance:

@@

For performance-critical applications, \SP{} includes a set of primitive types
that describe booleans and sized numbers.
Example types include \code{int64}, \code{uint64}, \code{double}.
There are also two special datatypes, \code{Array} and \code{Vector}, that store primitives.
The purpose of these types is to enable unboxed values and arithmetic at
runtime.
Consequently, their static semantics is a \emph{progressive}~\cite{pqk-onward-2012}
refinement over the semantics of basic Python types.

Primitive types are restricted in two ways to enforce a static separation
between fast-running primitives and standard data.
First, primitives cannot be implicitly cast to non-primitive types,
including the dynamic type.
This is another violation of the static gradual guarantee.
Second, operations such as \code{and} (boolean conjunction) do not accept a mix
of primitive and non-primitive data.

@@

To validate the design of \SP{}, we developed a model of the language in PLT
Redex~\cite{kcdeffmrtf-popl-2012}.
The model covers a substantial part of the Python language, including
assertions, loops, exception handlers, and delete statements.
It follows \SP{}'s approach to typing these features.
The model is available at:
\begin{center}
  \url{https://github.com/brownplt/insta-model}
\end{center}
Because there is no prior formalization of (all of) Python and of \SP{}, we cannot
\emph{verify} that the model matches them. Instead, we have applied
three methods to give confidence that our model matches reality.
First, we manually reviewed the model in depth---using our
own expertise---to look for non-conformance.  Second, 
we used Redex's random testing tools~\cite{kf-sfp-2009} to check
type soundness on thousands of examples (1,600 expressions and 11,000
programs).
% \footnote{Random testing helped us find and fix several issues in the
% model.  None of issues had implications for \SP{}'s soundness.}
Finally, we employed the well-established method of testing end-to-end
conformance with a test suite~\cite{gsk-ecoop-2010,gclpk-dls-2012,pmmwplck-oopsla-2013,bcfgmnss-popl-2014,fgpssmds-popl-2016}.
In particular, we translated \numSPtest{} tests from the \SP{} regression suite to the
syntax of the model and confirmed that the results
do match,
which suggests that the model conforms to actual \SP{}.

>>>>

%% TODO after submission, before camera-ready
%% - reword 'fields' to attribute/descriptor/property?
%% - 
%% - 
%% -

@@

Generic types are more difficult to enforce because a tag by itself does not
describe their contents.
For example, suppose that a function expects inputs that match
the type \code{Dict[int,int]}; that is, dictionaries with integer keys and
integer values.
A constant-time check for Python dictionaries is not enough to protect the
function against dicts with ill-typed elements.
The \SP{} solution is to provide a second kind of dictionary that ensures
the validity of its elements.
These different values have different types:
\begin{enumerate}
  \item
    The \emph{shallow} type $\sptrawpydict$ gets enforced with a tag check
    for dict values that ignores elements.
    Consequently, a parameterized surface type such as $\sptpydict{K}{V}$
    does not promise anything about keys and values.
  \item
    The \emph{concrete} type $\sptchkdict{K}{V}$ gets enforced with a
    three-part check for concrete dicts with exactly the same key and value
    types.
    The overall cost of such a check depends on the nesting of concrete types,
    but (crucially) does not depend on the size of incoming values.
    %% Programmers must use a type-annotated contructor to make a concrete-typed checked dict.
\end{enumerate}
%
These options keep run-time costs low and let programmers decide between
concrete and shallow types.
Often, the shallow types are more attractive because concrete types impose
nonlocal changes.
If one type changes to concrete, then several changes may need to happen:
all values that reach this type must be initialized as concrete,
all typed clients of these values must expect concrete values,
all values that reach those clients must be concrete,
and so on and so forth.

@@

\begin{table}[t]
  \newcommand{\nocost}{\textrm{---}}
  \caption{Costs of using shallow and concrete dictionaries in typed (T) and untyped (U) contexts relative to the number (N) of dict elements. A dash (\nocost{}) means zero cost.}
  \label{t:shallow-vs-concrete}

  \centering
  \(\begin{array}{l@{}rr@{\qquad}rr@{~~}@{\qquad}r@{~~}r}
                   & \mbox{Constructor} & \mbox{Cast} & \mbox{T-Read} & \mbox{T-Write} & \mbox{U-Read} & \mbox{U-Write} \\\hline
    \sptrawpydict         &   \nocost{} &        O(1) &          O(1) &      \nocost{} &     \nocost{} &      \nocost{} \\
    \sptchkdict{K}{V}     &        O(N) &        O(1) &     \nocost{} &      \nocost{} &     \nocost{} &           O(1)
  \end{array}\)
\end{table}

@@

Gradual class hierarchies are rarely studied in the literature, especially
for a language where truly untyped classes can enter the mix.
Thorn~\cite{wzlov-popl-2010} and SafeTS~\cite{rsfbv-popl-2015}, for example,
provide separate hierarchies for untyped and (gradually) typed classes.
The closest related work, for Nom~\cite{mt-oopsla-2017}, implements
a flexible design inspired by the gradual guarantees.
In particular, Nom subclasses can override any part of a signature with the
dynamic type.
\SP{} implements a simpler design that restricts types and
inheritance:

@@

For performance-critical applications, \SP{} includes a set of primitive types
that describe booleans and sized numbers.
Example types include \code{int64}, \code{uint64}, \code{double}.
There are also two special datatypes, \code{Array} and \code{Vector}, that store primitives.
The purpose of these types is to allow unboxed values and arithmetic at
runtime.
Consequently, their static semantics is a \emph{progressive}~\cite{pqk-onward-2012}
refinement over the semantics of basic Python types:
they permit fewer behaviors statically to enable the unboxing.
Operations such as \code{and} (boolean conjunction), for example, do not accept
a mix of primitive and non-primitive arguments.

Because performance is the top priority for primitive types, they
are intentionally incompatible with the dynamic type.
Casts between a primitive type and the dynamic type are rejected
statically---another violation of the gradual guarantee---to encourage the
spread of primitives in typed code.
At the boundaries to untyped Python code, however, \SP{} inserts casts.
These casts let programmers introduce primitive types in one module
without having to rewrite its untyped dependents.

@@

To validate the design of \SP{}, we developed a model of the language in PLT
Redex~\cite{kcdeffmrtf-popl-2012}.
The model covers a substantial part of the Python language, including
assertions, loops, exception handlers, and delete statements.
It follows \SP{}'s approach to typing these features.
The model is available at:
\begin{center}
  \url{https://github.com/brownplt/insta-model}
\end{center}
Because there is no prior formalization of (all of) Python and of \SP{}, we cannot
\emph{verify} that the model matches them. Instead, we have applied
three methods to give confidence that our model matches reality.
First, we manually reviewed the model in depth---using our
own expertise---to look for non-conformance.  Second, 
we used Redex's property-based testing tools~\cite{kf-sfp-2009} to check
type soundness on thousands of examples (1,600 expressions and 11,000
programs).
% \footnote{Property-based testing helped us find and fix several issues in the
% model.  None of issues had implications for \SP{}'s soundness.}
Finally, we employed the well-established method of testing end-to-end
conformance with a test suite~\cite{gsk-ecoop-2010,gclpk-dls-2012,pmmwplck-oopsla-2013,bcfgmnss-popl-2014,fgpssmds-popl-2016}.
In particular, we translated \numSPtest{} tests from the \SP{} regression suite to the
syntax of the model and confirmed that the results
do match,
which suggests that the model conforms to actual \SP{}.
